---
title: "finalproject"
output:
  html_document: default
  pdf_document: default
date: "2025-06-17"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## LOaD packages

```{r }
require(haven)
require(foreign)  # for loading spss data
require(sandwich)  # for robust standard errors
require(lme4)      # for multilevel analysis library(lmerTest)  # for posttests
require(lmtest)    # for robust standard errors and likelihood ratio tests
library(dplyr)
library(psych) # for descriptives
library(quest) # for descriptives of multi-level data
library(mlmhelpr) # for multi-level applications
library(ggplot2) # for plotting
require(stargazer) # for making nice tables
library(lmerTest) # for conducting post estimation test on ml models
library(performance) # for R2 calculation
library(lattice) # for plotting
library(plm) # for fixed effect models
library(tidyr) # tranforming data from wide to long and vice versa
library(broom.mixed) # for tidy() on mixed models
library(sjPlot)
pequiv <- readRDS("C:/Users/noaha/AppData/Local/Temp/fe870aea-e78e-46ec-9dd1-8757510678f3_SOEP-CORE.v39eu_R_EN.zip.8f3/R_EN/soepdata/pequiv.rds")
```

## Create Subset with relevant variables

```{r}
#pequiv$happiness <- pequiv$p11101
#pequiv$workhrs <- pequiv$e11101
#pequiv$employmentstatus <- pequiv$e11102
#pequiv$netincome <- pequiv$i11102
#pequiv$race <- pequiv$d11112ll
#pequiv$was_in_hospital <- pequiv$m11101
#pequiv$age <-pequiv$d11101
#d11109 years of education
#occupation e11105_v1
#m11126	Current Self-Rated Health Status
#x11101ll	Person Identification Number
#pid	Never Changing Person ID
#syear	Survey Year
#d11101	Age of Individual
#d11102ll	Gender of Individual
#cid	Original HH Number
#hid	Current Wave HH Number
#d11106	Number of Persons in HH
#d11107	Number of Children in HH
#d11104	Marital Status of Individual
#l11101	State of Residence
#e11103 employment level 

vars <- c(
  # Original core variables
  "d11109",        # Years of education
  "m11126",        # Self-rated health
  "pid",           # Person ID (stable)
  "syear",         # Year
  "d11101",        # Age
  "d11102ll",      # Gender
  "cid",           # Original household ID
  "hid",           # Current wave household ID
  "d11106",        # Household size
  "d11107",        # Number of children
  "d11104",        # Marital status
  "e11103",        #employment level
  "i11110",        #individual labor earnings
  "p11101",        # Happiness
  "e11102",        # Employment status
  "i11102",        # HH Net income
  "m11101"         # Was in hospital
)

#create subset
pequiv_subset <- pequiv[, vars]

#rename variables
pequiv_subset <- pequiv_subset %>%
  rename(
    education = d11109,
    health = m11126,
    pid = pid,
    year = syear,
    age = d11101,
    gender = d11102ll,
    hh_orig = cid,
    hh_wave = hid,
    hh_size = d11106,
    hh_kids = d11107,
    marital_status = d11104,
    happiness = p11101,
    employmentstatus = e11102,
    hhnetincome = i11102,
    was_in_hospital = m11101,
    employmentlvl = e11103,        #employment level
   indiv_wage =i11110       #individual labor earnings
  )

```

## Inspect Dataset
```{r}
str(pequiv_subset)

summary(pequiv_subset$happiness)
table(pequiv_subset$happiness)
summary(pequiv_subset$employmentstatus)
table(pequiv_subset$gender, useNA = "ifany")
table(pequiv_subset$occupation, useNA = "ifany")

attr(pequiv_subset$gender, "labels")

attr(pequiv_subset$marital_status, "labels")
```


## Data handling
Prepare data for analysis
```{r}
#all non-answers recoded to NA
pequiv_subset[] <- lapply(pequiv_subset, function(x) {
  if (is.numeric(x) || is.integer(x)) {
    x[x %in% -1:-8] <- NA
  }
  return(x)
})


#check
table(pequiv_subset$gender, useNA = "ifany")

#recode gender to new female variable with 0 male and 1 for female
pequiv_subset <- pequiv_subset %>%
  mutate(female = case_when(
    gender == 1 ~ 0,  # Male
    gender == 2 ~ 1,  # Female
    TRUE ~ NA_real_
  ))
table(pequiv_subset$female, useNA = "ifany")


pequiv_subset <- pequiv_subset %>%
  mutate(
    married = ifelse(marital_status == 1, 1, 0)
  )

#more preprocessing
#only core working age group
pequiv_subset <- pequiv_subset %>%
  filter(age >= 30, age <= 49)



#only beyond 2002 (less attrition more consistent survey questions)
pequiv_subset <- pequiv_subset %>%
  filter(year >= 2002, year <= 2020)
table(pequiv_subset$year, useNA = "ifany")


```
##more pre-processing
```{r}


#create income log
pequiv_subset$log_hhincome <- log1p(pequiv_subset$hhnetincome)




#create OECD equivalence scale for income
pequiv_subset <- pequiv_subset %>%
  mutate(eq_income = hhnetincome / sqrt(hh_size),
         log_eq_income = log1p(eq_income))  # if skewed





```

##check and handle missings
```{r}

colSums(is.na(pequiv_subset))
#significant missings:  health 18,000 ; married 18806 ; happiness 19397  ; hhnetincome 5721



# Visual check: does X differ by missingness in happiness?

# Income
ggplot(pequiv_subset, aes(x = is.na(happiness), y = log_eq_income)) +
  geom_boxplot() +
  labs(x = "Missing happiness?", y = "Log Household Income",
       title = "Does income differ between missing and non-missing happiness?")

# Employment status
ggplot(pequiv_subset, aes(x = factor(employmentstatus), fill = is.na(happiness))) +
  geom_bar(position = "fill") +
  labs(x = "Employment Status", y = "Proportion",
       fill = "Missing happiness?",
       title = "Does employment status relate to missing happiness?")

#health
ggplot(pequiv_subset, aes(x = is.na(happiness), y = health)) +
  geom_boxplot() +
  labs(x = "Missing happiness?", y = "Health",
       title = "Does health differ between missing and non-missing happiness?")

# Married
ggplot(pequiv_subset, aes(x = factor(married), fill = is.na(happiness))) +
  geom_bar(position = "fill") +
  labs(x = "Married", y = "Proportion",
       fill = "Missing happiness?",
       title = "Does marriage status relate to missing happiness?")

# Gender
ggplot(pequiv_subset, aes(x = factor(female), fill = is.na(happiness))) +
  geom_bar(position = "fill") +
  labs(x = "Gender", y = "Proportion",
       fill = "Missing happiness?",
       title = "Does gender relate to missing happiness?")



#remove incomplete observations on key variables
pequiv_subset <- pequiv_subset %>%
  filter(!is.na(health), !is.na(employmentstatus), !is.na(female), !is.na(happiness), !is.na(hhnetincome), !is.na(married))

```
##create income brackets and check 
```{r}

#create income per hh member brackets and calc mean happiness
pequiv_subset <- pequiv_subset %>%
  filter(!is.na(happiness), !is.na(hhnetincome)) %>%
  mutate(income_bracket2 = cut(
    eq_income,
    breaks = c(0, 15000, 30000, 45000, 70000, 120000, Inf),
    labels = c("1", "2", "3", "4", "5", "6"),
    right = FALSE  # bracket includes lower bound, excludes upper
  ))
mean_happiness_by_bracket <- pequiv_subset %>%
  group_by(income_bracket2) %>%
  summarise(
    n = n(),
    mean_happiness = mean(happiness, na.rm = TRUE)
  )

print(mean_happiness_by_bracket)
```


##check descriptives of variables of interest
```{r}
mean(pequiv_subset$happiness)
sd((pequiv_subset$happiness))
psych::describe(pequiv_subset$happiness)

table(pequiv_subset$employmentstatus)

```


## More data preprocessing
```{r}
#check number of ppl in dataset
length(unique(pequiv_subset$pid))

#check some correlations
cor(pequiv_subset$health, log1p(pequiv_subset$hhnetincome), use = "complete.obs")

cor(pequiv_subset$health, pequiv_subset$happiness, use = "complete.obs")
attr(pequiv_subset$health, "labels")
cor(pequiv_subset$health, pequiv_subset$employmentstatus, use = "complete.obs")
cor(pequiv_subset$happiness, pequiv_subset$employmentstatus, use = "complete.obs")
cor(pequiv_subset$happiness, pequiv_subset$hhnetincome, use = "complete.obs")
cor(pequiv_subset$health, pequiv_subset$age, use = "complete.obs")
cor(pequiv_subset$gender, pequiv_subset$hhnetincome, use = "complete.obs")
cor(pequiv_subset$happiness, pequiv_subset$year, use = "complete.obs")
cor(pequiv_subset$happiness, pequiv_subset$marital_status, use = "complete.obs")
attr(pequiv_subset$marital_status, "labels")
cor(pequiv_subset$happiness, log1p(pequiv_subset$hhnetincome), use = "complete.obs")

#try to plot log income over happiness
set.seed(123)  # for reproducibility

plot_data <- pequiv_subset %>%
  filter(!is.na(happiness), !is.na(hhnetincome)) %>%
  sample_n(5000)  # adjust sample size as needed

ggplot(plot_data, aes(x = log1p(hhnetincome), y = happiness)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", color = "darkgreen") +
  labs(
    title = "Happiness vs. log1p(Income)",
    x = "log(Income + 1)",
    y = "Happiness"
  )

```



#center and standardize some variables
```{r}

pequiv_subset <- pequiv_subset %>%
  mutate(
    # Center age around median
    age_c = age - median(age, na.rm = TRUE),
    
    # Time variable: years since 2002
    baseyear = year - 2002,
    
    # Standardized log household income
    log_eq_income_z = as.numeric(scale(log_eq_income)),
    
    # Number of children in household, centered around median
    hh_kids_c = hh_kids - median(hh_kids, na.rm = TRUE),
    
    # Household size centered around median
    hh_size_c = hh_size - median(hh_size, na.rm = TRUE),
    
    # Happiness centered around median
    happiness_c = happiness - median(happiness, na.rm = TRUE),
    
    # Health centered around median
    health_c = health - median(health, na.rm = TRUE)
    
    
  )

```
## create health lag
```{r}
# Create lagged health variable where the first observation per pid is retained
pequiv_subset <- pequiv_subset %>%
  arrange(pid, year) %>%
  group_by(pid) %>%
  mutate(
    health_lag = dplyr::lag(health),
    health_lag = if_else(row_number() == 1, health, health_lag)
  ) %>%
  ungroup()

# View side-by-side for a sample of individuals
pequiv_subset %>%
  filter(pid %in% sample(unique(pid), 10)) %>%
  dplyr::select(pid, year, health, health_lag) %>%
  arrange(pid, year)


```


###create descriptive table of key variables
```{r}

# key variables
desc_vars <- pequiv_subset %>%
  select(
    happiness,
    eq_income,
    log_eq_income,
    employmentstatus,
    female,
    married,
    baseyear,
    health_lag
  )

# Create descriptive statistics table
descriptives <- psych::describe(desc_vars)

# Print nicely for copy-paste
round(descriptives[, c("mean", "sd", "min", "max", "n")], 2)

```


#plot happiness by employmentstatus
```{r}
pequiv_subset %>%
  group_by(employmentstatus) %>%
  summarise(mean_happiness = mean(happiness, na.rm = TRUE),
            n = n())

happiness_summary <- pequiv_subset %>%
  filter(employmentstatus %in% c(0, 1)) %>%  # exclude NAs or other values
  group_by(employmentstatus) %>%
  summarise(avg_happiness = mean(happiness, na.rm = TRUE)) %>%
  mutate(
    employment_mean = if_else(employmentstatus == 1, "Employed", "Unemployed")
  )

pequiv_subset %>%
  filter(employmentstatus %in% c(0, 1)) %>%
  mutate(employment_mean = if_else(employmentstatus == 1, "Employed", "Unemployed")) %>%
  ggplot(aes(x = happiness, fill = employment_mean)) +
  geom_histogram(position = "identity", alpha = 0.6, bins = 20) +
  labs(title = "Distribution of Happiness by Employment Status",
       x = "Happiness",
       y = "Count",
       fill = "Employment Status") +
  theme_minimal() + 
  geom_vline(data = happiness_summary,
             aes(xintercept = avg_happiness, color = employment_mean),
             linetype = "dashed", size = 1)


# Calculate mean happiness by employment status
mean_happiness <- pequiv_subset %>%
  group_by(employmentstatus) %>%
  summarise(mean_happiness = mean(happiness, na.rm = TRUE))
mean_happiness$employmentstatus <- as.factor(mean_happiness$employmentstatus)

# Create the bar plot
ggplot(mean_happiness, aes(x = employmentstatus, y = mean_happiness, fill = employmentstatus)) +
  geom_bar(stat = "identity", width = 0.6) +
  scale_fill_manual(values = c("gray60", "gray80")) +
  labs(title = "Mean Life Satisfaction by Employment Status",
       x = "Employment Status",
       y = "Mean Life Satisfaction") +
  ylim(0, 10) +
  theme_minimal() +
  theme(
    legend.position = "none",
    text = element_text(size = 11)
  )
```


```{r}
hist(pequiv_subset$happiness)
psych::describe(pequiv_subset$happiness)

plot(pequiv_subset$hhnetincome)
plot(pequiv_subset$log_hhincome)
```






#simple regression and robust standard errors


#simple ols modes to compare

```{r}
ols1 <- lm(happiness ~ scale(log_eq_income) +  employmentstatus, data = pequiv_subset)
summary(ols1)

ols2 <- lm(happiness ~ scale(log_eq_income) + female*employmentstatus, data = pequiv_subset)
summary(ols2)

ols4 <- lm(happiness ~ scale(log_eq_income) + female*employmentstatus + scale(health_lag) + scale(baseyear) + married , data = pequiv_subset)
summary(ols4)


#ols2 <- lm(happiness ~ scale(log_hhincome) + relative_year +  relative_year2 +  female  + scale(health) + scale(baseyear) , data = pequiv_trans)
#summary(ols2)

psych::describe(pequiv_subset$hhnetincome)
quantile(pequiv_subset$hhnetincome, probs = seq(0, 1, 0.1), na.rm = TRUE)

pequiv_subset <- pequiv_subset %>%
  mutate(income_bracket_num = as.integer(income_bracket2) - 1)
table(pequiv_subset$income_bracket_num)

ols12 <- lm(happiness ~ scale(income_bracket_num) , data = pequiv_subset)
summary(ols12)

ols10 <- lm(happiness ~ scale(income_bracket_num) +  female*employmentstatus + scale(health_lag) + scale(baseyear) + married , data = pequiv_subset)
summary(ols10)

```
##create table with ols models and clustered se
```{r}
library(lmtest)
library(sandwich)
library(stargazer)

# Clustered standard errors by pid
cluster_se <- function(model, cluster_var) {
  vcovCL(model, cluster = cluster_var, type = "HC1")
}

# Compute clustered SEs
cl_ols1 <- cluster_se(ols1, ~pequiv_subset$pid)
cl_ols2 <- cluster_se(ols2, ~pequiv_subset$pid)
cl_ols4 <- cluster_se(ols4, ~pequiv_subset$pid)

# Stargazer with clustered SEs
stargazer(ols1, ols2, ols4,
          se = list(sqrt(diag(cl_ols1)), sqrt(diag(cl_ols2)), sqrt(diag(cl_ols4))),
          type = "text",  # or "text/ latex" / "html"
          title = "OLS Models Predicting Happiness with Clustered SEs",
          column.labels = c("OLS1", "OLS2", "OLS4"),
          dep.var.labels = "Happiness",
          no.space = TRUE,
          digits = 3,
          keep.stat = c("n", "rsq", "adj.rsq"))

```




##plot some random slopes
```{r}
library(dplyr)
library(lme4)

# Step 1: Sample 70 individuals
set.seed(123)
sampled_pids <- sample(unique(pequiv_subset$pid), 70)

# Step 2: Subset and clean data
d <- pequiv_subset %>%
  filter(pid %in% sampled_pids) %>%
  filter(!is.na(employmentstatus) & !is.na(happiness)) %>%
  mutate(employmentstatus = as.numeric(employmentstatus))

##  plot
ggplot(d, aes(x = employmentstatus, y = happiness, group = pid)) +
  geom_smooth(method = "lm", se = FALSE, aes(color = factor(pid))) +
  labs(x = "Employment Status (0 = Unemployed, 1 = Employed)",
       y = "Happiness",
       color = "Individual (pid)",
       title = "Individual OLS Regression Lines: Employment Status → Happiness") +
  theme_minimal() +
  theme(legend.position = "none")

```

#plot individual regression lines to see whether the slopes vary across participants

```{r}
# Step 3: Plot individual OLS regression lines
ggplot(d, aes(x = log_hhincome, y = happiness, group = pid)) +
  geom_point(alpha = 0.3, size = 0.8) +
  geom_smooth(method = "lm", se = FALSE, aes(color = factor(pid)), alpha = 0.5) +
  labs(x = "Logged Household Income",
       y = "Happiness",
       color = "Individual (pid)",
       title = "Individual OLS Regression Lines: Log Household Income → Happiness") +
  theme_minimal() +
  theme(legend.position = "none")
```


##1 MUlti-level model: NUll model, Level 1, Level 2, random slopes, cross level interaction
```{r}
#null model
mnull <- lmer(happiness ~ 1 + (1 | pid), data = pequiv_subset, REML = F)
summary(mnull)
#compute ICC
icc_null <- icc(mnull)
icc_null
#about 53 percent of the variance is at the pid level

#add key level 1 predictors
ml1a <- lmer(happiness ~  scale(log_eq_income)
                    + employmentstatus + (1 | pid),
                 data = pequiv_subset, , REML = F)
summary(ml1a)

#Add lvl1 predictors
ml1 <- lmer(happiness ~ scale(health_lag) + scale(log_eq_income) +  scale(baseyear) + married
                    + employmentstatus + (1 | pid),
                 data = pequiv_subset, , REML = F)
summary(ml1)
#add lvl2 predictors
ml2 <- lmer(happiness ~ scale(health_lag) + scale(log_eq_income) +  scale(baseyear) + married
                   + employmentstatus + female + (1 | pid),
                 data = pequiv_subset, REML = F)
summary(ml2)
#add random slopes
mrs <- lmer(happiness ~ scale(health_lag) + scale(log_eq_income) + scale(baseyear) + married
                   +employmentstatus + female + (1 +employmentstatus | pid),
                 data = pequiv_subset, REML = F)
summary(mrs)
#add cross-level interactions
m_interaction_rs <- lmer(happiness ~ scale(health_lag) + scale(log_eq_income) + scale(baseyear) + married
                               + employmentstatus * female +
                               (1 + employmentstatus | pid),
                             data = pequiv_subset,REML = F)

#add cross-level interaction 2
m_interaction2_rs <- lmer(happiness ~ scale(health_lag) + scale(log_eq_income)*female + scale(baseyear) + married
                               + employmentstatus * female +
                               (1 + employmentstatus | pid),
                             data = pequiv_subset,REML = F)
summary(m_interaction2_rs)
###this imodel is not in the paper since changes in effects are negligeble, and the additional explained variance so small that it would only make the table less readable


#check out final model
summary(m_interaction_rs)

#compute LRtests
lrtest(ml1, mnull)
lrtest(ml1a, mnull)
lrtest(ml1, ml1a)
lrtest(ml2, ml1)
lrtest(mrs, ml2)
lrtest(m_interaction_rs, mrs)
lrtest(m_interaction2_rs,m_interaction_rs)

#The likelihood ratio tests are statistically significant, implying that the
#more complex model fits the data statistically significantly better than
#the simpler model


#see whether random slope makes sense 
#yes: statistically significant
#ranova(m_interaction_rs)
```
#dotplot
```{r}
library(lattice)
dotplot_model<- dotplot(ranef(m_interaction_rs)) 
dotplot_model
```

## interpreting income effects
```{r}
# Mean and SD of log-income
mean_log_inc <- mean(pequiv_subset$log_eq_income, na.rm = TRUE)
sd_log_inc   <- sd(pequiv_subset$log_eq_income, na.rm = TRUE)

# A 1 SD increase means multiplying income by exp(sd_log_inc)

income_multiplier <- exp(sd_log_inc)
income_multiplier
#1SD increase in log income corresponds with 93% increase in income

#what happens when income doubles:

delta_log = log(2)        
delta_std = delta_log / sd_log_inc
delta_std
#Then multiply by your model coefficient:
delta_happiness = 0.12 * delta_std
delta_happiness
#doubling income means 0.126 higher expected happiness
```


##variance explained by interaction (Proportional Reduction in Random Slope Variance (PRV))
```{r}
# Full variance-covariance matrix
vc_mrs <- VarCorr(mrs)$pid
vc_int <- VarCorr(m_interaction_rs)$pid

# Extract random slope variances
var_slope_A <- as.numeric(vc_mrs[2, 2])
var_slope_B <- as.numeric(vc_int[2, 2])

# Calculate proportional reduction
prv_slope <- (var_slope_A - var_slope_B) / var_slope_A
prv_slope

((as.data.frame(VarCorr(mrs))[2,4])-(as.data.frame(VarCorr(m_interaction_rs))[2,4])) /
(as.data.frame(VarCorr(mrs))[2,4])
#1.4% of in between-person variance in the effect of employment on life satisfaction is explained by gender differences 

```



##Visualize
```{r}
library(emmeans)
# Step 1: Get estimated marginal means
emm1 <- emmeans(m_interaction_rs, ~ employmentstatus * female)

# Step 2: Convert to data frame
emm_df <- as.data.frame(emm1)
emm_df

# Step 3: Optional — relabel for clarity
emm_df <- emm_df %>%
  mutate(
    employment = factor(employmentstatus, labels = c("Unemployed", "Employed")),
    gender = factor(female, labels = c("Men", "Women"))
  )

# Step 4: Plot
employmentplot <- ggplot(emm_df, aes(x = employment, y = emmean, color = gender)) +
  geom_point(position = position_dodge(width = 0.15), size = 4) +  # thicker points
  geom_errorbar(aes(ymin = asymp.LCL, ymax = asymp.UCL),
                position = position_dodge(width = 0.15),
                width = 0.1, size = 1) +  # thicker CI bars
  labs(
    x = "Employment Status",
    y = "Predicted Life Satisfaction",
    color = "Gender",
    title = "Predicted Life Satisfaction by Employment Status and Gender"
  ) +
  theme_minimal(base_size = 10) +  # smaller overall sizing
  theme(
    plot.title = element_text(size = 12),
    axis.title = element_text(size = 10),
    axis.text = element_text(size = 9),
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9)
  )

employmentplot

```


##visualize life satisfaction for differing incomes
```{r}
##create a new grid with reasonable income values and predict life satisfaction for them
###try with four different gender/employment cmbinations

# Step 1: Create raw income range
income_vals <- seq(10000, 200000, by = 10000)
base_df <- expand.grid(
  raw_income = income_vals,
  employmentstatus = c(0, 1),
  female = c(0, 1)
)

# Step 2: Log1p-transform income
base_df$log_eq_income <- log1p(base_df$raw_income)

# Step 3: Standardize log_hhincome
mu <- mean(pequiv_subset$log_eq_income, na.rm = TRUE)
sigma <- sd(pequiv_subset$log_eq_income, na.rm = TRUE)
base_df$log_hhincome_scaled <- (base_df$log_eq_income - mu) / sigma

# Step 4: Fill in average values for other covariates
base_df <- base_df %>%
  mutate(
    health_lag = mean(pequiv_subset$health_lag, na.rm = TRUE),
    baseyear = mean(pequiv_subset$baseyear, na.rm = TRUE),
   married = mean(pequiv_subset$married, na.rm = TRUE)
  )
head(base_df)

# Step 5: Predict happiness (fixed effects only)
base_df$predicted_happiness <- predict(m_interaction_rs, newdata = base_df, re.form = NA)

# Step 6: Label groups for plotting
base_df <- base_df %>%
  mutate(
    group = case_when(
      employmentstatus == 1 & female == 0 ~ "Employed Men",
      employmentstatus == 1 & female == 1 ~ "Employed Women",
      employmentstatus == 0 & female == 0 ~ "Unemployed Men",
      employmentstatus == 0 & female == 1 ~ "Unemployed Women"
    )
  )

# Step 7: Plot
incomeplot <- ggplot(base_df, aes(x = raw_income, y = predicted_happiness, color = group)) +
  geom_line(size = 1.2) +
  labs(
    x = "Adjusted Household Income (in Euro)",
    y = "Predicted Life Satisfaction",
    color = "Group",
    title = "Predicted Life Satisfaction by Income, Gender, and Employment Status"
  ) +
  theme_minimal() +
  scale_color_manual(values = c(
    "Employed Men" = "#1f77b4",
    "Employed Women" = "#e377c2",
    "Unemployed Men" = "#2ca02c",
    "Unemployed Women" = "#d62728"
  ))
incomeplot

```



##checking explained variances
```{r}
# ml1a vs the null model
#explained var at lvl 1 by key predictors
1-((sigma(ml1a)^2/ (sigma(mnull)^2)))
#explained var at lvl 2
1-(as.numeric(VarCorr(ml1a)) / as.numeric(VarCorr(mnull)))


# ml1 vs the null model
#explained var at lvl 1
1-((sigma(ml1)^2/ (sigma(mnull)^2)))
#explained var at lvl 2
1-(as.numeric(VarCorr(ml1)) / as.numeric(VarCorr(mnull)))


#model_2 vs null model
#explained var at lvl 1
1-((sigma(ml2)^2/ (sigma(mnull)^2)))
#explained var at lvl 2
1-(as.numeric(VarCorr(ml2)) / as.numeric(VarCorr(mnull)))

```




```{r}


#different table like in hox book
#tab_model(
#  mnull, ml1, ml2, mrs, m_interaction_rs,
#  show.re.var = TRUE,       # Show random effects variances
#  show.icc = TRUE,          # Show ICC
#  show.aic = TRUE,          # Show AIC
#  show.bic = TRUE,          # Show BIC
#  dv.labels = c("M1: Null Model", "M2: L1 Predictors", "M3: +Gender", "M4: Random Slope", "M5: Interaction RS"),
#  title = "Multilevel Models Predicting Happiness"
#)




```


#MLM model assumptions
```{r}
model_interaction_rs <- m_interaction_rs

#residuals level 1
z_residuals_l1 <- resid(model_interaction_rs, type = "pearson", scale = T)

#save as DF
residuals_l1 <- data.frame(x = z_residuals_l1 )

hist(residuals_l1$x)
#not perectly normal distributed

qqmath(model_interaction_rs)
##deviations at the top and bottom



```
#Check whether the variance of the L1 residual errors is the same in all groups
```{r}
# Save the L1 residuals. let's add these as a variable to the dta
pequiv_subset$l1resid<-residuals(model_interaction_rs)

pequiv_subset$abs.l1resid<-abs(pequiv_subset$l1resid)
# absolute value of residuals
pequiv_subset$abs.l1resid2<-pequiv_subset$abs.l1resid^2
# squared absolute value of the residuals
Levene.Model.F <- lm(abs.l1resid2 ~ pid, data=pequiv_subset)
#ANOVA of the squared residuals
anova(Levene.Model.F) #displays the results

##significantly different at 5% f-test

```

```{r}
plot(model_interaction_rs, pid ~ resid(., scaled=TRUE))
##slide deviation to the left

plot(model_interaction_rs,
     form = resid(., type = "pearson") ~ fitted(.), abline=0)
##o check for failures of normality, nonlinearity, and heteroscedasticity. If these assumptions are met, the plotted points should be evenly divided above and below their mean value of zero. That is, there should be no relation between the variability around the zero-line and the fitted value (variability around the line should not be a function of the fitted values). 
plot(model_interaction_rs, type = c("p", "smooth"))
##smooth line is not horizontal but close enough
```

```{r}
#l2 data
l2_data <- pequiv_subset %>%
  group_by(pid) %>% # group data by clustering variable, student
  mutate(happiness_mean = mean(happiness)) %>% # create mean gpa per student
  select(pid, happiness_mean, female) %>%
  unique()


head(l2_data)


#residuals level 2
l2_intercept_resid = ranef(model_interaction_rs)$pid [, 1]
l2_slope_resid = ranef(model_interaction_rs)$pid[, 2]

#plot histograms
hist(l2_intercept_resid)

hist(l2_slope_resid)

#plot qq plots
qqmath(l2_intercept_resid)

qqmath(l2_slope_resid)
###level 2 residuals also not perfectly normal distributed but closer to normal distribution. affects standard errors


#add the lvl 2 residuals to lvl 2 data

l2_data$intercept_resid <- l2_intercept_resid
l2_data$slope_resid <- l2_slope_resid


# merge level-2 data to predicted scores (these predicted scores are at the
# l1-level)
pequiv_subset_pred <-pequiv_subset %>% inner_join(l2_data[,], by = "pid" ,
                                                  relationship="many-to-one")
pequiv_subset_pred

#add predicted scores
pequiv_subset_pred$predlmer = predict(model_interaction_rs)


pequiv_subset_pred %>%
  ggplot(mapping = aes(x = predlmer, y = intercept_resid)) +
  geom_point() +
  labs(x = "predicted value", y = "L2 residuals intercept")


pequiv_subset_pred %>%
  ggplot(mapping = aes(x = predlmer, y = slope_resid)) +
  geom_point() +
  labs(x = "predicted value", y = "L2 residuals slope")

# look at to figure out whether to standardize residuals
# https://link.springer.com/article/10.3758/s13428-021-01709-z#Sec34
#by Sara


l2_data %>%
  ggplot(mapping = aes(x = female, y = slope_resid)) +
  geom_point() +
  labs(x = "sex", y = "slope residuals")
##gender and slope residuals not strongly correlated
cor.test(l2_data$slope_resid, as.numeric(l2_data$female))
# gender not correlated to random slope (employmentstatus)


l2_data %>%
  ggplot(mapping = aes(x = intercept_resid, y = slope_resid)) +
  geom_point() +
  labs(x = "employment", y = "residuals")
#there seems to be a negative correlation between intercept and slope residuals
cor.test(l2_data$slope_resid, as.numeric(l2_data$intercept_resid))
#strong negative correlation. bad?

# by including a covariance between intercept and slope variance we
# account for a correlation between the intercept variance and slope variance! -> maybe need to do this?
```






# check for influential cases using leverage measures. use the function hlm_influence
```{r}
#doesnt work- too much CPU use
#infl <- hlm_influence(model_interaction_rs, level = "pid")
#print(infl, width = Inf)

```






##create variables to mark transitions to (un)-employment
```{r}
table(pequiv_subset$employmentstatus)
attr(pequiv_subset$employmentstatus, "labels")


pequiv_subset <- pequiv_subset %>%
  arrange(pid, year) %>%
  group_by(pid) %>%
  mutate(
    employmentstatus_lag = dplyr::lag(employmentstatus), 
    emp_to_unemp = if_else(employmentstatus_lag == 1 & employmentstatus == 0, 1, 0, missing = NA_integer_),
    unemp_to_emp = if_else(employmentstatus_lag == 0 & employmentstatus == 1, 1, 0, missing = NA_integer_)
  ) %>%
  ungroup()

# Summary tables
table(pequiv_subset$emp_to_unemp, useNA = "always")
table(pequiv_subset$unemp_to_emp, useNA = "always")
```

#filter dataset for those who become unemployed
```{r}
ids_emp_to_unemp <- pequiv_subset %>%
  filter(emp_to_unemp == 1) %>%
  distinct(pid)

employment_df <- pequiv_subset %>%
  filter(pid %in% ids_emp_to_unemp$pid)

length(unique(employment_df$pid))


```

##plot happiness time to transitioning from emp to unemp
```{r}

# Step 1: Identify transitions from employment to unemployment
transition_years <- pequiv_subset %>%
  filter(emp_to_unemp == 1) %>%
  select(pid, transition_year = year)

# Step 2: Identify control group (never transitioned)
control_ids <- pequiv_subset %>%
  anti_join(transition_years, by = "pid") %>%
  distinct(pid)

# Step 3: Assign "fake" transition year to control group (median year per person)
control_fake_transition <- pequiv_subset %>%
  filter(pid %in% control_ids$pid) %>%
  group_by(pid) %>%
  summarise(transition_year = floor(median(year, na.rm = TRUE)))

# Step 4: Create treated group with relative year
treated_traj <- pequiv_subset %>%
  inner_join(transition_years, by = "pid") %>%
  mutate(
    relative_year = year - transition_year,
    group = "Transitioned"
  ) %>%
  filter(relative_year %in% -5:5)

# Step 5: Create control group with relative year
control_traj <- pequiv_subset %>%
  inner_join(control_fake_transition, by = "pid") %>%
  mutate(
    relative_year = year - transition_year,
    group = "Control"
  ) %>%
  filter(relative_year %in% -5:5)

# Step 6: Combine and summarize with confidence intervals
happiness_compare <- bind_rows(treated_traj, control_traj) %>%
  group_by(group, relative_year) %>%
  summarise(
    mean_happiness = mean(happiness, na.rm = TRUE),
    sd = sd(happiness, na.rm = TRUE),
    n = sum(!is.na(happiness)),
    se = sd / sqrt(n),
    ci_lower = mean_happiness - 1.96 * se,
    ci_upper = mean_happiness + 1.96 * se,
    .groups = "drop"
  )

# Step 7: Plot
ggplot(happiness_compare, aes(x = relative_year, y = mean_happiness, color = group, fill = group)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper), alpha = 0.2, color = NA) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  labs(
    title = "Average Happiness Around Employment → Unemployment Transition",
    x = "Years Relative to Transition",
    y = "Average Happiness",
    color = "Group",
    fill = "Group"
  ) +
  theme_minimal()



```

```{r}
# Summarize transition group happiness trajectory with CIs
happiness_transitions_only <- treated_traj %>%
  group_by(relative_year) %>%
  summarise(
    mean_happiness = mean(happiness, na.rm = TRUE),
    sd = sd(happiness, na.rm = TRUE),
    n = sum(!is.na(happiness)),
    se = sd / sqrt(n),
    ci_lower = mean_happiness - 1.96 * se,
    ci_upper = mean_happiness + 1.96 * se,
    .groups = "drop"
  )

# Plot
ggplot(happiness_transitions_only, aes(x = relative_year, y = mean_happiness)) +
  geom_line(color = "steelblue", size = 1.2) +
  geom_point(color = "steelblue", size = 2) +
  geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper), alpha = 0.2, fill = "steelblue") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  labs(
    title = "Average Happiness Around Employment → Unemployment Transition",
    x = "Years Relative to Transition",
    y = "Average Happiness"
  ) +
  theme_minimal()

```
##separately for men and women
```{r}
# Step 1: Label gender using `female` variable
treated_traj <- treated_traj %>%
  mutate(
    gender_label = if_else(female == 1, "Women", "Men")
  )

# Step 2: Summarize happiness by gender and relative year
happiness_by_gender <- treated_traj %>%
  group_by(gender_label, relative_year) %>%
  summarise(
    mean_happiness = mean(happiness, na.rm = TRUE),
    sd = sd(happiness, na.rm = TRUE),
    n = sum(!is.na(happiness)),
    se = sd / sqrt(n),
    ci_lower = mean_happiness - 1.96 * se,
    ci_upper = mean_happiness + 1.96 * se,
    .groups = "drop"
  )

# Step 3: Plot
ggplot(happiness_by_gender, aes(x = relative_year, y = mean_happiness, color = gender_label, fill = gender_label)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper), alpha = 0.2, color = NA) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  labs(
    title = "Happiness Around Employment → Unemployment Transition, by Gender",
    x = "Years Relative to Transition",
    y = "Average Happiness",
    color = "Gender",
    fill = "Gender"
  ) +
  theme_minimal()


```






